---
title: "CTmax analysis"
author: "Hannah von Hammerstein and Sonia Bejarano"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "85%"
)

set.seed(2026)
dir.create(here::here("figures"), showWarnings = FALSE)
dir.create(here::here("results"), showWarnings = FALSE)
```

# Overview

This document reproduces the CTmax analyses from the cleaned dataset supplied
with this repository. The primary inference uses the first experimental ramp
and compares the three acclimation temperatures (28, 30, and 32 °C). The
second ramp, available for the 28 and 32 °C treatments, is used to assess
repeatability.

# Packages and data

```{r packages-and-data}
library(dplyr)
library(emmeans)
library(ggplot2)
library(readxl)

data_path <- here::here("Data_CTmaxHist.xlsx")
stopifnot(file.exists(data_path))

ctmax_data <- read_excel(data_path, sheet = "CTmaxHistology")

required_columns <- c(
  "ramp_num", "Temp_Treat", "CTMax", "Histopathology", "C.Temp_Treat"
)
stopifnot(all(required_columns %in% names(ctmax_data)))

temperature_levels <- c("a.ambient.28", "b.medium.30", "c.hot.32")

ctmax_data <- ctmax_data %>%
  mutate(
    ramp_num = factor(ramp_num, levels = c("first", "second")),
    C.Temp_Treat = as.numeric(C.Temp_Treat),
    Temperature_C = factor(C.Temp_Treat, levels = c(28, 30, 32)),
    Temp_Treat = factor(
      case_when(
        C.Temp_Treat == 28 ~ "a.ambient.28",
        C.Temp_Treat == 30 ~ "b.medium.30",
        C.Temp_Treat == 32 ~ "c.hot.32",
        TRUE ~ NA_character_
      ),
      levels = temperature_levels
    )
  )

dplyr::glimpse(ctmax_data)
```

# Primary analysis: Ramp 1

```{r primary-model}
ramp1_data <- ctmax_data %>%
  filter(ramp_num == "first") %>%
  droplevels()

stopifnot(nlevels(droplevels(ramp1_data$Temperature_C)) == 3)
ctmax_model <- lm(CTMax ~ Temperature_C, data = ramp1_data)

summary(ctmax_model)
anova(ctmax_model)
```

## Model checks

The following diagnostic figures show residuals against fitted values,
residual distributions by acclimation temperature, and the residual
normal Q–Q plot. These are retained because they support the model checks
reported in the manuscript.

```{r model-diagnostic-figures, fig.width=10, fig.height=3.5, out.width='100%'}
par(mfrow = c(1, 3))

plot(
  fitted(ctmax_model),
  residuals(ctmax_model),
  xlab = "Fitted values",
  ylab = "Residuals",
  pch = 16
)
abline(h = 0, lty = 2)

boxplot(
  residuals(ctmax_model) ~ ramp1_data$Temperature_C,
  xlab = "Temperature treatment",
  ylab = "Residuals"
)

qqnorm(residuals(ctmax_model), pch = 16)
qqline(residuals(ctmax_model), col = "red")

par(mfrow = c(1, 1))
```

```{r model-diagnostic-tests}
car::leveneTest(CTMax ~ Temperature_C, data = ramp1_data)
shapiro.test(residuals(ctmax_model))
kruskal.test(CTMax ~ Temperature_C, data = ramp1_data)
```

## Treatment estimates

```{r treatment-estimates}
ctmax_counts <- ramp1_data %>%
  group_by(Temperature_C, C.Temp_Treat) %>%
  summarise(
    n = n(),
    observed_mean = mean(CTMax, na.rm = TRUE),
    observed_sd = sd(CTMax, na.rm = TRUE),
    .groups = "drop"
  )

ctmax_emmeans <- emmeans(ctmax_model, ~ Temperature_C)
ctmax_emmeans_df <- as.data.frame(ctmax_emmeans) %>%
  mutate(
    C.Temp_Treat = c(28, 30, 32)
  )

ctmax_counts
ctmax_emmeans_df
```

# Repeatability: Ramp 1 versus Ramp 2

The second ramp contains observations from the 28 and 32 °C treatments. Welch
two-sample t-tests compare CTmax between ramps separately within those two
treatments.

```{r ramp-comparisons}
ramp_comparison_data <- ctmax_data %>%
  filter(Temp_Treat %in% c("a.ambient.28", "c.hot.32")) %>%
  droplevels()

ramp_summary <- ramp_comparison_data %>%
  group_by(C.Temp_Treat, ramp_num) %>%
  summarise(
    n = sum(!is.na(CTMax)),
    mean_CTmax = mean(CTMax, na.rm = TRUE),
    sd_CTmax = sd(CTMax, na.rm = TRUE),
    .groups = "drop"
  )

ramp_tests <- lapply(
  split(ramp_comparison_data, ramp_comparison_data$C.Temp_Treat),
  function(x) t.test(CTMax ~ ramp_num, data = x, var.equal = FALSE)
)

ramp_summary
ramp_tests
```

# Acclimation response ratio

The acclimation response ratio (ARR) is the model-estimated change in CTmax
divided by the corresponding change in acclimation temperature:

\[
ARR = \frac{\Delta CTmax}{\Delta T_{acclimation}}
\]

```{r arr-analysis}
arr_contrasts <- contrast(
  ctmax_emmeans,
  method = list(
    "30_vs_28" = c(-1, 1, 0),
    "32_vs_28" = c(-1, 0, 1),
    "32_vs_30" = c(0, -1, 1)
  )
)

arr_results <- confint(arr_contrasts, adjust = "none") %>%
  as.data.frame() %>%
  mutate(
    comparison = recode(
      contrast,
      "30_vs_28" = "30 C vs 28 C",
      "32_vs_28" = "32 C vs 28 C",
      "32_vs_30" = "32 C vs 30 C"
    ),
    delta_Tacc_C = recode(
      contrast,
      "30_vs_28" = 2,
      "32_vs_28" = 4,
      "32_vs_30" = 2
    ),
    delta_CTmax_C = estimate,
    lower_delta_CTmax_C = lower.CL,
    upper_delta_CTmax_C = upper.CL,
    ARR = delta_CTmax_C / delta_Tacc_C,
    ARR_lower_95_CI = lower_delta_CTmax_C / delta_Tacc_C,
    ARR_upper_95_CI = upper_delta_CTmax_C / delta_Tacc_C
  ) %>%
  dplyr::select(
    comparison,
    delta_Tacc_C,
    delta_CTmax_C,
    lower_delta_CTmax_C,
    upper_delta_CTmax_C,
    ARR,
    ARR_lower_95_CI,
    ARR_upper_95_CI
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

arr_results

write.csv(
  arr_results,
  here::here("results", "CTmax_ARR_results.csv"),
  row.names = FALSE
)
```

# Final CTmax figure

This is the only CTmax figure printed in the rendered analysis. Circles and
solid estimates represent Ramp 1; triangles and dashed estimates represent
Ramp 2. Ramp 2 was conducted only at 28 and 32 °C.

```{r final-ctmax-figure, fig.width=7, fig.height=5, fig.cap='CTmax by acclimation temperature and experimental ramp.'}
ramp2_data <- ctmax_data %>%
  filter(ramp_num == "second") %>%
  droplevels()

ramp2_model <- lm(CTMax ~ Temperature_C, data = ramp2_data)

ramp2_emmeans_df <- emmeans(ramp2_model, ~ Temperature_C) %>%
  as.data.frame() %>%
  mutate(C.Temp_Treat = c(28, 32))

temp_fill <- c(
  "a.ambient.28" = "#ffe79a",
  "b.medium.30" = "#f8c0bb",
  "c.hot.32" = "#ee7063"
)

temp_colour <- c(
  "a.ambient.28" = "#E6C600",
  "b.medium.30" = "#E18E8B",
  "c.hot.32" = "#C85B52"
)

p_ctmax_final <- ggplot() +
  geom_violin(
    data = filter(ctmax_data, ramp_num == "second"),
    aes(C.Temp_Treat, CTMax, fill = Temp_Treat, colour = Temp_Treat),
    trim = FALSE,
    bw = 0.05,
    alpha = 0.2,
    scale = "width",
    linetype = "dashed",
    show.legend = FALSE
  ) +
  geom_jitter(
    data = filter(ctmax_data, ramp_num == "second"),
    aes(C.Temp_Treat, CTMax, fill = Temp_Treat),
    shape = 24,
    size = 4.5,
    colour = "black",
    width = 0.2
  ) +
  geom_violin(
    data = filter(ctmax_data, ramp_num == "first"),
    aes(C.Temp_Treat, CTMax, fill = Temp_Treat, colour = Temp_Treat),
    trim = FALSE,
    bw = 0.05,
    alpha = 0.7,
    scale = "width",
    show.legend = FALSE
  ) +
  geom_jitter(
    data = filter(ctmax_data, ramp_num == "first"),
    aes(C.Temp_Treat, CTMax, fill = Temp_Treat),
    shape = 21,
    size = 4.5,
    colour = "black",
    alpha = 0.7,
    width = 0.2
  ) +
  geom_errorbar(
    data = ramp2_emmeans_df,
    aes(C.Temp_Treat, ymin = lower.CL, ymax = upper.CL),
    width = 0.30,
    linewidth = 1.3,
    linetype = "longdash",
    colour = "grey40",
    position = position_nudge(x = 0.06)
  ) +
  geom_point(
    data = ramp2_emmeans_df,
    aes(C.Temp_Treat, emmean),
    shape = 24,
    size = 6.8,
    stroke = 1.15,
    alpha = 0.9,
    fill = "grey40",
    colour = "grey40",
    position = position_nudge(x = 0.06)
  ) +
  geom_errorbar(
    data = ctmax_emmeans_df,
    aes(C.Temp_Treat, ymin = lower.CL, ymax = upper.CL),
    width = 0.30,
    linewidth = 1.3,
    colour = "grey20",
    position = position_nudge(x = -0.06)
  ) +
  geom_point(
    data = ctmax_emmeans_df,
    aes(C.Temp_Treat, emmean),
    shape = 21,
    size = 7,
    stroke = 1.15,
    fill = "grey20",
    colour = "grey20",
    position = position_nudge(x = -0.06)
  ) +
  scale_x_continuous(breaks = c(28, 30, 32)) +
  scale_fill_manual(values = temp_fill) +
  scale_colour_manual(values = temp_colour) +
  labs(
    x = "Acclimation temperature (°C)",
    y = "CTmax (°C)"
  ) +
  theme_bw(base_size = 20) +
  theme(
    panel.border = element_rect(linetype = "solid", fill = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "black", size = 18),
    axis.ticks = element_line(linewidth = 1, colour = "black"),
    axis.ticks.length = grid::unit(0.2, "cm"),
    axis.title = element_text(face = "bold", size = 20),
    plot.caption = element_text(size = 14, hjust = 0.5, face = "italic"),
    legend.position = "none"
  )

p_ctmax_final

ggsave(
  here::here("figures", "CTmax_combined_ramps.pdf"),
  p_ctmax_final,
  width = 4.5,
  height = 3.5,
  units = "in",
  scale = 2.5
)

ggsave(
  here::here("figures", "CTmax_combined_ramps.png"),
  p_ctmax_final,
  width = 4.5,
  height = 3.5,
  units = "in",
  scale = 2.5,
  dpi = 600
)
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
