---
title: "Histopathology analysis"
author: "Hannah von Hammerstein and Sonia Bejarano"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "85%"
)

set.seed(2026)
dir.create(here::here("figures"), showWarnings = FALSE)
dir.create(here::here("results"), showWarnings = FALSE)
```

# Overview

This document reproduces the histopathology analyses from the cleaned dataset
supplied with this repository. Histopathology is coded as a binary response
(0 = no alteration detected; 1 = alteration detected). The primary analysis
uses the first experimental ramp and tests whether alteration probability
differs among the 28, 30, and 32 °C acclimation treatments. The second ramp,
available at 28 and 32 °C, is used to assess repeatability.

# Packages and data

```{r packages-and-data}
library(dplyr)
library(emmeans)
library(ggplot2)
library(readxl)

data_path <- here::here("Data_CTmaxHist.xlsx")
stopifnot(file.exists(data_path))

histo_data <- read_excel(data_path, sheet = "CTmaxHistology")

required_columns <- c(
  "ramp_num", "Temp_Treat", "Histopathology", "C.Temp_Treat"
)
stopifnot(all(required_columns %in% names(histo_data)))

temperature_levels <- c("a.ambient.28", "b.medium.30", "c.hot.32")

histo_data <- histo_data %>%
  mutate(
    ramp_num = factor(ramp_num, levels = c("first", "second")),
    C.Temp_Treat = as.numeric(C.Temp_Treat),
    Temperature_C = factor(C.Temp_Treat, levels = c(28, 30, 32)),
    Temp_Treat = factor(
      case_when(
        C.Temp_Treat == 28 ~ "a.ambient.28",
        C.Temp_Treat == 30 ~ "b.medium.30",
        C.Temp_Treat == 32 ~ "c.hot.32",
        TRUE ~ NA_character_
      ),
      levels = temperature_levels
    ),
    Histopathology = as.integer(Histopathology)
  )

stopifnot(all(na.omit(histo_data$Histopathology) %in% c(0L, 1L)))
dplyr::glimpse(histo_data)
```

# Primary analysis: Ramp 1

A binomial logistic regression estimates histopathological alteration
probability as a function of categorical acclimation temperature.

```{r primary-model}
ramp1_data <- histo_data %>%
  filter(ramp_num == "first") %>%
  droplevels()

stopifnot(nlevels(droplevels(ramp1_data$Temperature_C)) == 3)

histo_model <- glm(
  Histopathology ~ Temperature_C,
  data = ramp1_data,
  family = binomial(link = "logit")
)

histo_null_model <- glm(
  Histopathology ~ 1,
  data = ramp1_data,
  family = binomial(link = "logit")
)

summary(histo_model)
```

## Model comparison and checks

The likelihood-ratio test compares the temperature model with an intercept-only
model. The residual deviance divided by residual degrees of freedom is reported
as a simple overdispersion diagnostic.

```{r model-comparison}
aic_table <- AIC(histo_null_model, histo_model)
aic_table$delta_AIC <- aic_table$AIC - min(aic_table$AIC)

likelihood_ratio_test <- anova(
  histo_null_model,
  histo_model,
  test = "Chisq"
)

dispersion_ratio <- deviance(histo_model) / df.residual(histo_model)

aic_table
likelihood_ratio_test
dispersion_ratio
```

The diagnostic plot is retained because it supports the model assessment
reported in the manuscript.

```{r model-diagnostic-figure, fig.width=7, fig.height=5}
plot(
  fitted(histo_model),
  residuals(histo_model, type = "deviance"),
  xlab = "Fitted probability",
  ylab = "Deviance residual",
  pch = 16
)
abline(h = 0, lty = 2)
```

## Observed and model-estimated probabilities

```{r probability-estimates}
histo_counts_ramp1 <- ramp1_data %>%
  group_by(Temperature_C, C.Temp_Treat) %>%
  summarise(
    n = n(),
    alterations = sum(Histopathology, na.rm = TRUE),
    observed_probability = mean(Histopathology, na.rm = TRUE),
    .groups = "drop"
  )

histo_emmeans_ramp1 <- emmeans(
  histo_model,
  ~ Temperature_C,
  type = "response"
)

histo_probabilities_ramp1 <- as.data.frame(histo_emmeans_ramp1) %>%
  mutate(C.Temp_Treat = c(28, 30, 32)) %>%
  rename(
    predicted_probability = prob,
    lower_95_CI = asymp.LCL,
    upper_95_CI = asymp.UCL
  ) %>%
  left_join(histo_counts_ramp1, by = c("Temperature_C", "C.Temp_Treat")) %>%
  arrange(C.Temp_Treat) %>%
  mutate(
    across(
      c(observed_probability, predicted_probability, lower_95_CI, upper_95_CI),
      ~ round(.x, 3)
    )
  ) %>%
  dplyr::select(
    Temperature_C,
    C.Temp_Treat,
    n,
    alterations,
    observed_probability,
    predicted_probability,
    lower_95_CI,
    upper_95_CI
  )

histo_probabilities_ramp1

write.csv(
  histo_probabilities_ramp1,
  here::here("results", "Histopathology_Ramp1_probabilities.csv"),
  row.names = FALSE
)
```

## Odds ratios

```{r odds-ratios}
odds_ratios <- data.frame(
  term = names(coef(histo_model)),
  odds_ratio = exp(coef(histo_model)),
  exp(confint(histo_model)),
  check.names = FALSE
)

odds_ratios
```

# Repeatability: Ramp 1 versus Ramp 2

Fisher's exact tests compare the frequency of histopathological alterations
between ramps separately at 28 and 32 °C.

```{r ramp-comparisons}
ramp_comparison_data <- histo_data %>%
  filter(Temp_Treat %in% c("a.ambient.28", "c.hot.32")) %>%
  droplevels()

ramp_tables <- lapply(
  split(ramp_comparison_data, ramp_comparison_data$C.Temp_Treat),
  function(x) table(ramp = x$ramp_num, alteration = x$Histopathology)
)

ramp_fisher_tests <- lapply(ramp_tables, fisher.test)

ramp_tables
ramp_fisher_tests
```

# Ramp 2 model

The second-ramp model is a sensitivity analysis restricted to the two
temperatures represented in that ramp.

```{r ramp2-model}
ramp2_data <- histo_data %>%
  filter(ramp_num == "second") %>%
  droplevels()

histo_model_ramp2 <- glm(
  Histopathology ~ Temperature_C,
  data = ramp2_data,
  family = binomial(link = "logit")
)

histo_null_model_ramp2 <- glm(
  Histopathology ~ 1,
  data = ramp2_data,
  family = binomial(link = "logit")
)

anova(histo_null_model_ramp2, histo_model_ramp2, test = "Chisq")
summary(histo_model_ramp2)
```

# Final histopathology figure

This is the only histopathology figure printed in the rendered analysis.
Raw binary outcomes are shown at 0 (absent) and 1 (present). Circles and solid
estimates represent Ramp 1; triangles and dashed estimates represent Ramp 2.
No line is drawn between categorical treatment estimates.

```{r final-histopathology-figure, fig.width=7, fig.height=5, fig.cap='Histopathological alteration outcomes and model-estimated probabilities by acclimation temperature and ramp.'}
histo_emmeans_ramp1_df <- as.data.frame(histo_emmeans_ramp1) %>%
  mutate(C.Temp_Treat = c(28, 30, 32))

histo_emmeans_ramp2_df <- emmeans(
  histo_model_ramp2,
  ~ Temperature_C,
  type = "response"
) %>%
  as.data.frame() %>%
  mutate(
    C.Temp_Treat = c(28, 32),
    x_plot = C.Temp_Treat + 0.22
  )

plot_data <- histo_data %>%
  mutate(
    x_plot = C.Temp_Treat + case_when(
      Temp_Treat == "b.medium.30" ~ 0,
      ramp_num == "first" ~ -0.16,
      ramp_num == "second" ~ 0.16,
      TRUE ~ 0
    )
  )

temp_fill <- c(
  "a.ambient.28" = "#ffe79a",
  "b.medium.30" = "#f8c0bb",
  "c.hot.32" = "#ee7063"
)

temp_colour <- c(
  "a.ambient.28" = "#E6C600",
  "b.medium.30" = "#E18E8B",
  "c.hot.32" = "#C85B52"
)

p_histo_final <- ggplot() +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = -0.075,
    ymax = 0.075,
    fill = "grey94",
    colour = NA
  ) +
  annotate(
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = 0.925,
    ymax = 1.075,
    fill = "grey94",
    colour = NA
  ) +
  geom_point(
    data = plot_data,
    aes(x_plot, Histopathology, fill = Temp_Treat, shape = ramp_num),
    position = position_jitter(width = 0.13, height = 0.022, seed = 2026),
    size = 4.5,
    colour = "black",
    alpha = 0.7
  ) +
  geom_errorbar(
    data = histo_emmeans_ramp2_df,
    aes(x_plot, ymin = asymp.LCL, ymax = asymp.UCL),
    width = 0.30,
    linewidth = 1.3,
    linetype = "dashed",
    colour = "grey40"
  ) +
  geom_point(
    data = histo_emmeans_ramp2_df,
    aes(x_plot, prob),
    shape = 24,
    size = 6.8,
    stroke = 1,
    alpha = 0.9,
    fill = "grey40",
    colour = "grey40"
  ) +
  geom_errorbar(
    data = histo_emmeans_ramp1_df,
    aes(C.Temp_Treat, ymin = asymp.LCL, ymax = asymp.UCL),
    width = 0.30,
    linewidth = 1.3,
    colour = "grey20"
  ) +
  geom_point(
    data = histo_emmeans_ramp1_df,
    aes(C.Temp_Treat, prob),
    shape = 21,
    size = 7,
    stroke = 1.15,
    fill = "grey20",
    colour = "grey20"
  ) +
  scale_x_continuous(
    breaks = c(28, 30, 32),
    expand = expansion(mult = c(0.08, 0.10))
  ) +
  scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = c("Absent", "0.25", "0.50", "0.75", "Present"),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_manual(values = temp_fill, guide = "none") +
  scale_colour_manual(values = temp_colour, guide = "none") +
  scale_shape_manual(
    values = c("first" = 21, "second" = 24),
    labels = c("Ramp 1", "Ramp 2"),
    name = NULL
  ) +
  coord_cartesian(
    xlim = c(27.55, 32.55),
    ylim = c(-0.075, 1.075),
    clip = "off"
  ) +
  labs(
    x = "Acclimation temperature (°C)",
    y = "Probability of histopathological alteration"
  ) +
  guides(
    shape = guide_legend(
      override.aes = list(
        fill = "grey75",
        colour = "grey25",
        alpha = 1,
        size = 3.2
      )
    )
  ) +
  theme_bw(base_size = 20) +
  theme(
    panel.border = element_rect(linetype = "solid", fill = NA),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "black", size = 18),
    axis.ticks = element_line(linewidth = 1, colour = "black"),
    axis.ticks.length = grid::unit(0.2, "cm"),
    axis.title = element_text(face = "bold", size = 20),
    axis.title.x = element_text(margin = margin(t = 15)),
    legend.position = "none"
  )

p_histo_final

ggsave(
  here::here("figures", "Histopathology_combined_ramps.pdf"),
  p_histo_final,
  width = 4.5,
  height = 3.3,
  units = "in",
  scale = 2.5
)

ggsave(
  here::here("figures", "Histopathology_combined_ramps.png"),
  p_histo_final,
  width = 4.5,
  height = 3.3,
  units = "in",
  scale = 2.5,
  dpi = 600
)
```

# Reproducibility information

```{r session-information}
sessionInfo()
```
